---
title: "tutorial_v1_1"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{tutorial_v1_1}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# **Untargeted Metabolomics/Lipidomics Analysis**

## **Contents:**
## 1. Analysis Setup
### 1.1. Check Dependencies and Environment
### 1.2. Define Parameters and Upload Data
### 1.3. Upload Data Files

## 2. Data Integrity Check
### 2.1. Quality Control **(Devlopment)**
### 2.2. Data Transformation, Scaling, Normalization, Imputation
### 2.3. Annotation Summary

## 3. Descriptive Statistics
### 3.1. Fold Change Calculation
### 3.2. Group Class Sums/Class Ratios **(Devlopment)**

## 4. Statistics and Hypothesis Testing
### 4.1. Dimension Reduction: PCA/UMAP/PLSDA
### 4.2. Statistics: ANOVA 1-, 2-, or 3-Variable and Visualization **(Devlopment)**
### 4.3. ChemRICH and Visualization
### 4.4. Lipid Network Plots **(Development)**
### 4.5. Correlation **(Development)**
### 4.6. Regression **(Devlopment)**
### 4.7. Network Analysis **(Devlopment)**
### 4.8. Tanglegram/ML-based comparisons between data sets **(Devlopment)**

## **Scripts:**
## 1. Analysis Setup
```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  dpi = 500,
  fig.width = 7
)
```

### 1.1. Check Dependencies and Environment
```{r, echo = TRUE, fig.align= "center", results=F, message=F, warning=F, eval = FALSE}
# Check if library is installed
if("Stevens.MSAnalyze" %in% unique(installed.packages()[,1]) == FALSE){ #nolint
  devtools::install_github("cschasestevens/Stevens.MSAnalyze",
    ref = "master",
    build_vignettes = TRUE
  )
}
# Check dependencies
library(Stevens.MSAnalyze)
ms_check_lib()
# Set working directory
getwd()
browseVignettes("Stevens.MSAnalyze")
# Connect to local database (only if performing ChemRICH for lipidomics data)
## Verify that lipidref.db has been downloaded and placed in the "ref" folder
## from the "main" branch of Stevens.MSAnalyze before connecting to database
setwd("/home/ncsteven/analyze/Chase.MSAnalyze")
db1 <- DBI::dbConnect(RSQLite::SQLite(), "ref/lipidref.db")
## Use DBI::dbDisconnect(db1) followed by
## DBI::unlink("database name") to disconnect from the database
```

### 1.2. Define Parameters and Upload Data
```{r, echo = TRUE, fig.align= "center", results=F, message=F, warning=F, eval = FALSE}
# Load data or use example data
d <- ms_input(
  # Metadata column number
  md_num = 4,
  # Have the data been normalized?
  norm1 = TRUE,
  # Generate a QC report?
  qc_rep = FALSE
)

# Add annotation information from reference
d[["anno"]] <- d[["anno"]][d[["anno"]][["type"]] == "an.comp", ]
d[["anno"]] <- cbind(
  d[["anno"]],
  d[["anno.ref"]]
)
```

## 2. Data Check
### 2.1. Quality Control **(Devlopment)**
```{r, echo = TRUE, fig.align= "center", results=F, message=F, warning=F, eval = FALSE}

```

### 2.2. Data Transformation, Scaling, Normalization, and Imputation 
```{r, echo = TRUE, fig.align= "center", results=F, message=F, warning=F, eval = FALSE}
# Check data
d <- ms_data_check(
  ld = d,
  cl_var = "Group",
  samp_id = "label"
)
# Plot distribution
ggplot2::ggsave(
  "analysis/plot.dist.png",
  d[["plots"]][["plot.dist"]],
  width = 21,
  height = 7,
  dpi = 300
)
# Plot sample correlation
png(
  "analysis/plot.heatmap.sample.correlation.org.png",
  width = 32,
  height = 26,
  units = "cm",
  res = 1000
)
print(d[["plots"]][["plot.cor"]])
dev.off()
```

### 2.3. Annotation Summary
```{r, echo = TRUE,fig.align="center",results=F,message=F,warning=F,eval = FALSE}
# Total and Group-wise Annotation Information (Displayed as Sunburst plot)
names(d[["data"]][["anno"]])
d_an <- ms_plot_anno(
  ld = d,
  md = c("Major_Class", "Class", "Subclass")
)
# Plot annotations
ggplot2::ggsave(
  "analysis/plot.anno.png",
  d_an[["plot"]],
  width = 16,
  height = 14,
  dpi = 600
)
```

## 3. Descriptive Statistics
### 3.1. Fold Change Calculation
```{r, echo = TRUE, fig.align= "center", results=F, message=F, warning=F, eval = FALSE}
# Fold Change - Individual
names(d[["data"]])
d[["fc"]] <- ms_stat_fc(
  mat1 = d[["data"]][["imputed"]],
  md = d[["data"]][["meta"]],
  md_var = c("Group"),
  an = d[["data"]][["anno"]]
)
head(d[["fc"]])
## Fold Change - Class
d[["fcc"]] <- ms_stat_fc(
  mat1 = d[["data"]][["imputed"]],
  md = d[["data"]][["meta"]],
  md_var = c("Group"),
  an = d[["data"]][["anno"]],
  fc_class = TRUE,
  grp_class = "Subclass"
)
head(d[["fcc"]])
```

### 3.2 Group Class Means/Class Ratios **(Development)**
```{r, echo = TRUE, results=F, message=F, warning=F, eval = FALSE}

```

## 4. Statistics and Hypothesis Testing
### 4.1. Dimension Reduction: PCA/UMAP/PLSDA
```{r, echo = TRUE, fig.align= "center", results=F, message=F, warning=F, eval = FALSE}
library(dplyr)
set.seed(1234)
# PCA
d[["plots"]][["plot.pca"]] <- ms_dim_rd(
  mat1 = d[["data"]][["data.pareto"]],
  md = d[["data"]][["meta"]],
  md_var = c("Group"),
  dim_type = "PCA",
  dim1 = "2D",
  p_type = "single"
)
ggplot2::ggsave(
  "analysis/plot.dr2D.pca.Group.png",
  d[["plots"]][["plot.pca"]],
  width = 16,
  height = 8,
  dpi = 600
)
# UMAP
d[["plots"]][["plot.umap"]] <- ms_dim_rd(
  mat1 = d[["data"]][["data.pareto"]],
  md = d[["data"]][["meta"]],
  md_var = c("Group"),
  dim_type = "UMAP",
  dim1 = "3D",
  p_type = "single"
)
ggplot2::ggsave(
  "analysis/plot.dr2D.umap.Group.png",
  d[["plots"]][["plot.umap"]],
  width = 16,
  height = 8,
  dpi = 600
)
```

### 4.2. Statistics: ANOVA 1-, 2-, or 3-Variable and Visualization
```{r, echo = TRUE, results=F, message=F, warning=F, eval = FALSE}
# Run Stats
d[["stats"]] <- ms_stat_anova(
  matpv = d[["data"]][["data.pareto"]],
  matfc = d[["data"]][["imputed"]],
  md = d[["data"]][["meta"]],
  md_var = c("Group"),
  an = d[["data"]][["anno"]]
)
# Determine number of significant compounds in each comparison
num_sig <- setNames(
  dplyr::count(
    d[["stats"]][d[["stats"]][["FDR"]] < 0.05, ],
    d[["stats"]][d[["stats"]][["FDR"]] < 0.05, c("Comparison.fc")]
  ),
  c("Comparison", "n")
)
nrow(num_sig)
unique(num_sig[["Comparison"]])
## Save
write.table(
  num_sig,
  "analysis/table.numsig.txt",
  sep = "\t",
  col.names = TRUE,
  row.names = FALSE
)
write.table(
  d[["stats"]],
  "analysis/table.stats.txt",
  col.names = TRUE,
  row.names = FALSE,
  sep = "\t"
)

# Plots (Volcano, Heatmap, and Violin Plot **(Development)**)
## Volcano Plot
### Single
p1 <- ms_plot_volcano(
  d_stat = d[["stats"]],
  comp_name = "comp1-comp2",
  diff_col = "Log2FC",
  p_col = "p adj",
  p_cut = 0.05,
  f_cut = 0.25,
  f_lim = 4,
  y_limit = 20,
  x_title = paste("Increased in comp1 <-----> Increased in comp2")
)
ggplot2::ggsave(
  paste(
    "analysis/plot.vol.",
    "comp1.vs.comp2",
    ".png",
    sep = ""
  ),
  p1,
  width = 16,
  height = 10,
  dpi = 400
)
### All (for comparisons with at least one significantly altered compound)
lapply(
  seq.int(1, length(unique(num_sig[["Comparison"]])), 1),
  function(x) {
    tryCatch(
      {
        ggplot2::ggsave(
          paste(
            "analysis/plot.vol.",
            gsub("-", ".vs.", unique(num_sig[["Comparison"]])[[x]]),
            ".png",
            sep = ""
          ),
          ms_plot_volcano(
            d_stat = d[["stats"]],
            comp_name = unique(num_sig[["Comparison"]])[[x]],
            diff_col = "Log2FC",
            p_col = "p adj",
            p_cut = 0.05,
            f_cut = 0.25,
            f_lim = 4,
            y_limit = 20,
            x_title = paste(
              "Increased in",
              strsplit(unique(num_sig[["Comparison"]])[[x]], "-")[[1]][[2]],
              "<----->",
              " Increased in",
              strsplit(unique(num_sig[["Comparison"]])[[x]], "-")[[1]][[1]],
              sep = " "
            )
          ),
          width = 16,
          height = 10,
          dpi = 400
        )
      },
      error = function(e) {
        print("plot generation unsuccessful, skipping comparison...")
      }
    )
  }
)

## Heatmaps
### Between region
h_out <- ms_plot_heat(
  d_stat = d[["stats"]],
  d_ref = d[["data"]][["anno"]],
  c_list = c(
    "comp1a-comp2a", "comp1b-comp2b"
  ),
  diff_col = "Log2FC",
  p_col = "FDR",
  an1 = c("Major.Class", "Saturation"),
  hm_w = 36,
  hm_h = 10,
  fs_r = 8,
  cl_c = TRUE,
  cl_r = TRUE
)
png(
  "analysis/plot.heatmap.comp1.vs.comp2.png",
  width = 44,
  height = 12,
  units = "cm",
  res = 1000
)
print(h_out)
dev.off()
```

### 4.3. ChemRICH and Visualization
```{r, echo = TRUE, fig.align= "center", results=F, message=F, warning=F, eval = FALSE}
# Run ChemRICH
d[["chemrich"]] <- ms_stat_crich(
  d_stat = d[["stats"]],
  d_ref = d[["data"]][["anno"]],
  cl_name = "label_saturation"
)
head(d[["chemrich"]])

# Return comparisons with at least one significantly enriched cluster
cr_sig <- d[["chemrich"]][d[["chemrich"]][["FDR"]] < 0.05, ]
head(cr_sig)
length(unique(cr_sig[["Comparison"]]))
nrow(cr_sig)
unique(cr_sig[["Cluster"]])
unique(cr_sig[["label"]])

# Plot ChemRICH with significantly enriched clusters
lapply(
  seq.int(1, length(unique(cr_sig[["Comparison"]])), 1),
  function(x) {
    p1 <- ms_plot_crich(
      dfcr = d[["chemrich"]][
        d[["chemrich"]][["Comparison"]] == unique(cr_sig[["Comparison"]])[[x]],
      ]
    )
    ggplot2::ggsave(
      paste(
        "analysis/plot.cr.",
        gsub("-", "_", unique(cr_sig[["Comparison"]])[[x]]),
        ".png",
        sep = ""
      ),
      p1,
      width = 16,
      height = 10,
      dpi = 400
    )
  }
)
```

### 4.4 Lipid Network Plots **(Development)**
```{r, echo = TRUE, results=F, message=F, warning=F, eval = FALSE}
head(d[["chemrich"]])
## Pathway Legend
ms_plot_lipidnet(
  net_type = "pathways_legend"
)
## Base network
ms_plot_lipidnet(
  net_type = "base"
)
## Plot dataset annotations
ms_plot_lipidnet(
  d_enrch = d[["chemrich"]],
  net_type = "annotation",
  cl_var = "label_saturation"
)
## Plot dataset enrichment results
ms_plot_lipidnet(
  d_enrch = d[["chemrich"]],
  net_type = "enrichment",
  cl_var = "label_saturation",
  comp1 = "HFA_F-BR",
  comp_sat = TRUE
)
## Save
ggplot2::ggsave(
  "analysis/testnet.png",
  ms_plot_lipidnet(
    d_enrch = d[["chemrich"]],
    net_type = "enrichment",
    cl_var = "label_saturation",
    comp1 = "HFA_F-BR",
    comp_sat = TRUE
  ),
  width = 16,
  height = 12,
  dpi = 600
)
## Save legend only
ggplot2::ggsave(
  "analysis/testnet_legend.png",
  ms_plot_lipidnet(
    net_type = "pathways_legend"
  ),
  width = 8,
  height = 6,
  dpi = 600
)
```